Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-05-08 manuscript.Rmd
Updates revision

#### $\hat{\bm{R}}$
Next to a visual inspection of convergence, a numerical inspection is usually performed as well. A common approach is to check $\hat{R}$, also called the Gelman-Rubin Diagnostic [@gelman1992inference]. $\hat{R}$ is the ratio of between-chain variance and within-chain variance [@sorensen2015bayesian]. If the chains diverge, the between-chain variance will be higher than the within-chain variance, resulting in a $\hat{R}$ greater than 1. This indicates that the chains have not converged. There are different types of $\hat{R}$, such as the split- and rank-based $\hat{R}$ [@gelman2013bayesian; @vehtari2019rank]. Even though researchers tend to trust convergence statistics more than visual inspection, these statistics are also not fool-proof. Therefore, the development of better convergence statistics is a current topic in Bayesian analysis methods.

The $\hat{R}$ values for all individual and general symbolic distance effect parameters computed by `rstan` and `brms` are shown in panel C of Figure \@ref(fig:traceplots3p). The figure shows that $\hat{R}$'s are not substantially greater than 1. Therefore, it seems that the distributions have converged.

#### Number of Effective Samples 
Another popular convergence statistic is the number of effective samples. This statistic is concerned with the dependence of posterior samples. In MCMC algorithms, the samples are to some degree dependent on one another: the parameter values at iteration *i* are similar to the parameter values at iteration *i-1*. The number of effective samples represents the estimated total number of *independent* draws from the posterior for every model parameter [@stan2018b, section 15.4]. If the number of effective samples is low, this indicates a high dependency of iterations in the sampling procedure. Therefore, the number of effective samples can be seen as a measure of the amount of new information about the posterior distribution that is provided by the total number of samples drawn from the posterior. A common rule for interpretation is that the number should be close to or even greater than the number of total iterations [@vehtari2019rank; @neff2018].

>>>>

#### $\hat{\bm{R}}$
Next to a visual inspection of convergence, a numerical inspection is usually performed as well. A common approach is to check $\hat{R}$, also called the Gelman-Rubin Diagnostic [@gelman1992inference]. $\hat{R}$ is the ratio of between-chain variance and within-chain variance [@sorensen2015bayesian]. If the chains diverge, the between-chain variance will be higher than the within-chain variance, resulting in a $\hat{R}$ greater than 1. This indicates that the chains have not converged. \Revision{The current adhered criteria is that $\hat{R}$ should not exceed 1.01 }[@vehtari2019rank]. There are different types of $\hat{R}$, such as the split- and rank-based $\hat{R}$ [@gelman2013bayesian; @vehtari2019rank]. Even though researchers tend to trust convergence statistics more than visual inspection, these statistics are also not fool-proof. \Revision{For instance, in hierarchical modeling, with many parameters, it is more likely that parameters will exceed the threshold while there are no issues with the parameter estimation. It is, therefore, advisable to evaluate parameters carefully.} The development of better convergence statistics is a current topic in Bayesian analysis methods.

The $\hat{R}$ values for all individual and general symbolic distance effect parameters computed by `rstan` and `brms` are shown in panel C of Figure \@ref(fig:traceplots3p). The figure shows that $\hat{R}$'s are not substantially greater than \Revision{1.01}. Therefore, it seems that the distributions have converged.

#### Number of Effective Samples 
Another popular convergence statistic is the number of effective samples (ESS). This statistic is concerned with the dependence of posterior samples. In MCMC algorithms, the samples are to some degree dependent on one another: the parameter values at iteration *i* are similar to the parameter values at iteration *i-1*. The number of effective samples represents the estimated total number of *independent* draws from the posterior for every model parameter [@stan2018b, section 15.4]. If the number of effective samples is low, this indicates a high dependency of iterations in the sampling procedure. Therefore, the number of effective samples can be seen as a measure of the amount of new information about the posterior distribution that is provided by the total number of samples drawn from the posterior. \Revision{A common rule for interpretation is that the number of effective samples should equal at least 100 per chain} [@vehtari2019rank]. \Revision{In the case of four chains, this would mean a value of 400 or higher. The higher the value the better. There are different ways to estimate the number of effective samples. Bulk-ESS looks, for instance, at number of effective samples at the centre of the posterior distribution, while Tail-ESS evaluates the tails of the posterior distribution (). In this tutorial, ... .}

### 2023-05-08 manuscript.tex
Updates revision

\begin{equation}
\begin{aligned}
\mu_{\gamma} &\sim \text{Student's t}(3, 1, 10), \label{eq:chosenpriors1} \\ 
\mu_{\beta}, \mu_{\delta} &\sim \text{Normal}(0, 1), \\ 
\sigma, \sigma_{\gamma}, \sigma_{\beta}, \sigma_{\delta} &\sim \text{Truncated-Student's t}_{+}(3, 0, 10). 
\end{aligned}
\end{equation}
The prior distributions \texttt{brms} specifies on the standard deviations can be transformed to prior distributions on variances (\(var\)). Prior distributions on variances have a smaller scale than priors on standard deviation (\(sd\)). Note, however, that the transformation of the prior distributions is not as simple as the transformation of the parameter itself. \textcolor{blue}{If we formulate a prior on standard deviations, there is always an implied prior on the variances as well. If we think of the distribution as $f(x)$, $x$ can either be defined as $\sqrt{(var)}$, if we formulated the prior on the standard deviation, or $x$ can be defined as $(sd^2)$ if we formulated the prior on the variance. From this, we can derive the implied distribution $f(var)$ or $f(sd)$. Notably, this is not the same as calculating $\sqrt{(f(x))}$ or $f(x)^2$. To transform the priors on standard deviations to variances, we try to find the parameters that make the prior distribution match the implied distribution.} In our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/R\%20code/Prior\%20transformation}{Online Supplement}, we provide R-code with a numeric solution to the transformation. For the \texttt{brms} priors, we obtain the following implied prior distributions on variances:

@@

Next to a visual inspection of convergence, a numerical inspection is usually performed as well. A common approach is to check \(\hat{R}\), also called the Gelman-Rubin Diagnostic (Gelman \& Rubin, 1992). \(\hat{R}\) is the ratio of between-chain variance and within-chain variance (Sorensen \& Vasishth, 2015). If the chains diverge, the between-chain variance will be higher than the within-chain variance, resulting in a \(\hat{R}\) greater than 1. This indicates that the chains have not converged. There are different types of \(\hat{R}\), such as the split- and rank-based \(\hat{R}\) (Gelman et al., 2013; Vehtari, Gelman, Simpson, Carpenter, \& Bürkner, 2019). Even though researchers tend to trust convergence statistics more than visual inspection, these statistics are also not fool-proof. Therefore, the development of better convergence statistics is a current topic in Bayesian analysis methods.

The \(\hat{R}\) values for all individual and general symbolic distance effect parameters computed by \texttt{rstan} and \texttt{brms} are shown in panel C of Figure \ref{fig:traceplots3p}. The figure shows that \(\hat{R}\)'s are not substantially greater than 1. Therefore, it seems that the distributions have converged.

@@

Another popular convergence statistic is the number of effective samples. This statistic is concerned with the dependence of posterior samples. In MCMC algorithms, the samples are to some degree dependent on one another: the parameter values at iteration \emph{i} are similar to the parameter values at iteration \emph{i-1}. The number of effective samples represents the estimated total number of \emph{independent} draws from the posterior for every model parameter (Stan Development Team, 2018a, sec. 15.4). If the number of effective samples is low, this indicates a high dependency of iterations in the sampling procedure. Therefore, the number of effective samples can be seen as a measure of the amount of new information about the posterior distribution that is provided by the total number of samples drawn from the posterior. A common rule for interpretation is that the number should be close to or even greater than the number of total iterations (Carpenter, 2018; Vehtari, Gelman, Simpson, Carpenter, \& Bürkner, 2019).

@@

\leavevmode\hypertarget{ref-neff2018}{}%
Carpenter, B. (2018). We were measuring the speed of stan incorrectly -- it's faster than we thought in some cases due to antithetical sampling {[}blog post{]}. Retrieved from \url{https://statmodeling.stat.columbia.edu/2018/01/18/\%20measuring-speed-stan-incorrectly-faster-thought-cases-due-antithetical-sampling/}

>>>>

\begin{equation}
\begin{aligned}
\mu_{\gamma} &\sim \text{Student's t}(3, 1, 10), \label{eq:chosenpriors1} \\ 
\mu_{\beta}, \mu_{\delta} &\sim \text{Normal}(0, 1), \\ 
\sigma, \sigma_{\gamma}, \sigma_{\beta}, \sigma_{\delta} &\sim \text{Truncated-Student's t}_{+}(3, 0, 10). 
\end{aligned}
\end{equation}
The prior distributions \texttt{brms} specifies on the standard deviations can be transformed to prior distributions on variances (\(var\)). Prior distributions on variances have a smaller scale than priors on standard deviation (\(sd\)). Note, however, that the transformation of the prior distributions is not as simple as the transformation of the parameter itself. \textcolor{blue}{If we formulate a prior on standard deviations, there is always an implied prior on the variances as well. If we think of the distribution as $f(x)$, $x$ can either be defined as $\sqrt{(var)}$, if we formulated the prior on the standard deviation, or $x$ can be defined as ${sd}^2$ if we formulated the prior on the variance. From this, we can derive the implied distribution $f(var)$ or $f(sd)$. Notably, this is not the same as calculating $\sqrt{(f(x))}$ or $f(x)^2$. To transform the priors on standard deviations to variances, we try to find the parameters that match the prior distribution to the implied distribution.} In our \href{https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/R\%20code/Prior\%20transformation}{Online Supplement}, we provide R-code with a numeric solution to the transformation. For the \texttt{brms} priors, we obtain the following implied prior distributions on variances:

@@

Next to a visual inspection of convergence, a numerical inspection is usually performed as well. A common approach is to check \(\hat{R}\), also called the Gelman-Rubin Diagnostic (Gelman \& Rubin, 1992). \(\hat{R}\) is the ratio of between-chain variance and within-chain variance (Sorensen \& Vasishth, 2015). If the chains diverge, the between-chain variance will be higher than the within-chain variance, resulting in a \(\hat{R}\) greater than 1. This indicates that the chains have not converged. \textcolor{blue}{The current adhered criteria is that $\hat{R}$ should not exceed 1.01 }(Vehtari, Gelman, Simpson, Carpenter, \& Bürkner, 2019). There are different types of \(\hat{R}\), such as the split- and rank-based \(\hat{R}\) (Gelman et al., 2013; Vehtari, Gelman, Simpson, Carpenter, \& Bürkner, 2019). Even though researchers tend to trust convergence statistics more than visual inspection, these statistics are also not fool-proof. \textcolor{blue}{For instance, in hierarchical modeling, with many parameters, it is more likely that parameters will exceed the threshold while there are no issues with the parameter estimation. It is, therefore, advisable to evaluate separate parameters carefully.} The development of better convergence statistics is a current topic in Bayesian analysis methods.

The \(\hat{R}\) values for all individual and general symbolic distance effect parameters computed by \texttt{rstan} and \texttt{brms} are shown in panel C of Figure \ref{fig:traceplots3p}. The figure shows that \(\hat{R}\)'s are not substantially greater than \textcolor{blue}{1.01}. Therefore, it seems that the distributions have converged.

@@

Another popular convergence statistic is the number of effective samples (ESS). This statistic is concerned with the dependence of posterior samples. In MCMC algorithms, the samples are to some degree dependent on one another: the parameter values at iteration \emph{i} are similar to the parameter values at iteration \emph{i-1}. The number of effective samples represents the estimated total number of \emph{independent} draws from the posterior for every model parameter (Stan Development Team, 2018a, sec. 15.4). If the number of effective samples is low, this indicates a high dependency of iterations in the sampling procedure. Therefore, the number of effective samples can be seen as a measure of the amount of new information about the posterior distribution that is provided by the total number of samples drawn from the posterior. \textcolor{blue}{A common rule for interpretation is that the number should be at least 100 per chain.}(Vehtari, Gelman, Simpson, Carpenter, \& Bürkner, 2019) \textcolor{blue}{In case of four chains, this would mean a value of 400 or higher. The higher the value the better. There are different ways to estimate the number of effective samples. Bulk-ESS looks, for instance, at number of effective samples at the centre of the posterior distribution, while Tail-ESS evaluates the tails of the posterior distribution (). In this tutorial, ... .}

@@
